---
title: "Portfolio Optimization in R"
output:
        prettydoc::html_pretty:
                theme: architect
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd('/Users/COLLABERA TECH/Downloads/upwork/')
```

This document presents the steps in conducting portfolio optimization in R. GitHub repository is found [here](https://github.com/imjbmkz/portfolio_optimization).

<br><br>

### Importing packages required

We will use several packages that will help us pull stocks data and to visualize risks and returns. The following code installs the packages required if they're not yet installed. Afterwards, they will be loaded in our current R session.

```{r install_packages, warning=F, message=F}
## (1) Define the packages that will be needed
packages <- c('tidyquant', 'quantmod', 'TTR', 'PerformanceAnalytics', 'PortfolioAnalytics')

## (2) Install them if not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

## (3) Load the packages into R session
invisible(lapply(packages, library, character.only = TRUE))
```

<br><br>

### Getting stocks adjusted prices data 

Let's now load the stocks data that we will be needing. First, we will define a variable called `myStocks` that has the stock symbols that we want. Modify these based on what you need.

```{r define_stock_codes}
## Create a character vector that has the stock codes we need
myStocks <- c('AUB', 'EW', 'PSB', 'PTC', 'UBP')
```

Then, let's define the script that will load these stocks into R. We use `lapply` to "*apply*" a function across a list or a vector. The below code is going through all stock codes in `myStocks` variable, and loads it using the `getSymbols` function from quantmod package. Note that we are getting the weekly stocks data, and the parameter `auto.assign` is set to `FALSE` to load the `xts` object into `myStocks` variable. Finally, we give the list index names based on the data that they represent.

```{r, warning=F, message=F}
## Load the stocks needed into R 
myStocks <- lapply(myStocks, function(x) {getSymbols(
        x, periodicity='weekly', auto.assign=FALSE)})

## Set the names in the list to easily access them.
names(myStocks) <- c('AUB', 'EW', 'PSB', 'PTC', 'UBP')
```

We can look at the first few row of the stocks data that we loaded.

```{r head_stocks}
## View first three rows of selected stock
head(myStocks$AUB, 3)
```

For this exercise, we are only interested in the adjusted prices of the stocks. The `Ad` function still from quantmod package does that. We again use `lapply` to use `Ad` function across all other stocks. To better work with the data, this has been transformed into an `xts` object instead of just a `list`.

```{r get_adj_price}
## Get adjusted prices of all stocks
myStocks_ad <- lapply(myStocks, Ad)

## Transform into xts
myStocks_ad <- do.call(merge, myStocks_ad)
```

Let's see how the data looks like.

```{r}
## View first 3 rows of the combined different adjusted prices 
head(myStocks_ad, 3)
```

<br><br>

### Getting weekly log returns 

Now that we have our merged adjusted prices of different stock codes, let's get their corresponding weekly returns. We still use `lapply`, `do.call`, and `merge` functions to do it. We are removing the first row since there are no returns on the first observation in the data.

```{r get_weekly_log_ret}
## Make a list that contains log weekly returns of each stock
myStocks_returns <- lapply(myStocks_ad, weeklyReturn, type='log')

## Tranform into an xts object
myStocks_returns <- do.call(merge, myStocks_returns)

## Adjust the column names 
colnames(myStocks_returns) <- c(
        'AUB.weekly.returns', 'EW.weekly.returns', 'PSB.weekly.returns',
        'PTC.weekly.returns', 'UBP.weekly.returns')

## Remove first row since these do not have returns
myStocks_returns <- myStocks_returns[-1]
```

Let's look at it.

```{r}
head(myStocks_returns, 3)
```

To avoid several steps above, we can use `Return.calculate` function from PerformanceAnalytics package. We still need to remove the first row as it pertains to the first observation.

```{r performance_analytics_returns}
## Use Return.calculate to easily get returns, and remove -1
myStocks_returns_PA <- Return.calculate(myStocks_ad, method='log')[-1]
```

Let's look at it.

```{r}
head(myStocks_returns_PA, 3)
```

Now that we're done calculating weekly returns of each stock, let's get the returns for the entire portfolio. We're generating random weights for each stock, then using those weights, we calculate our returns. $PortfolioReturns = \sum{stocks * weights}$

```{r random_weights}
## Generate random weights
random_weights <- runif(5)
random_weights <- random_weights / sum(random_weights)

## Generate portfolio weighted returns
portfolio_return <- Return.portfolio(myStocks_returns, random_weights)

## Generate portfolio weighted returns with rebalancing
portfolio_return_rebalance <- Return.portfolio(myStocks_returns, random_weights, rebalance_on='months')

## Combine results 
portfolio_return_both <- cbind(portfolio_return, portfolio_return_rebalance)

## Change column names
colnames(portfolio_return_both) <- c('non_rebalanced', 'monthly_rebalanced')
```

Let's also get the annualized returns of the data with risk-free rate of 10% divided by 52 weeks. This is added on the `Rf` parameter of `table.AnnualizedReturns` function.

```{r}
annual_returns_table <- table.AnnualizedReturns(portfolio_return_both, Rf=0.1/52)
annual_returns_table
```

Based on the above results, rebalancing the data gave a lower return, but a slightly lower risk than the non_rebalanced weights.

Let's now proceed to portfolio optimization.

<br><br> 

### Portfolio optimization

<br>

#### Step 1: Define the specification
We will define a portfolio specification object using `portfolio.spec` function from PortfolioPerformance package. This will serve as our placeholder so that we can feed other metrics, such as constraints, into this object.
```{r portfolio_specs}
port_spec <- portfolio.spec(assets=colnames(myStocks_returns))
```

<br>

#### Step 2: Adding the constraints required
We will add constraints that will control the values that will be assigned to each parameter. Specifically, we want: 

1. The weighted sum to be exactly 100%, but we're adding a small interval of 0.01, so weighted sum can be between 0.99 to 1.01. 

2. We want the weights to be positive. It must not be a negative number.

3. Finally, we want to control the minimum and maximum weights that will be assigned to individual stock.

```{r constraints}
## Define first constraint
port_spec <- add.constraint(portfolio=port_spec, type='weight_sum', min_sum=0.99, max_sum=1.01)

## Define second constraint
port_spec <- add.constraint(portfolio=port_spec, type='long_only')

## Define third constraint
port_spec <- add.constraint(portfolio=port_spec, type='box', min=0.1, max=0.3)
```

<br>

#### Step 3: Add the objective of the portfolio
In portfolio optimization, the objective is to either minimize the risk or to maximize returns. Our objective is to minimize the risk The following code will define that objective.

```{r minimize_risk}
port_spec_min <- add.objective(portfolio=port_spec, type='risk', name='StdDev')
```

We'll now start optimizing the portfolio to minimize the risk.

<br>

#### Step 4: Optimization - minimizing risk
We will now start with the optimization process to minimize risk. We make another variable called `optimize_min_risk`. We set the randomization seed to make sure that in the next run of this markdown file, we still get the same results.

```{r optimize_risk, warning=F, message=F}
set.seed(1234)
optimize_min_risk <- optimize.portfolio(
        R=myStocks_returns, portfolio=port_spec_min, 
        optimize_method='random')
```

Afterwards, we can get the best weights per asset. 

```{r}
optimize_min_risk$weights
```

The above plot tells us that in order to get a minimum risk, the weights of each asset must be as defined above. We can also get the minimized risk from the model, and compare it with the individual risk associated per asset.

```{r}
## Get optimized risk
min_risk <- optimize_min_risk$objective_measures$StdDev[1,1]

## Get risk per asset
asset_risks <- sapply(myStocks_returns, sd)

## Check if optimized risk is lower than individual risks
for (i in 1:5) {
        if (min_risk < asset_risks[[i]]) {
                cat('Optimized risk is lower than ', 
                    names(asset_risks)[i], '\' risk\n', sep='')
        } else {
                cat('Optimized risk is higher than ', 
                    names(asset_risks)[i], '\' risk\n', sep='')
        }
}
```

We can see from the above results that our optimized risk is lower than the individual risks of assets.






